---
title: "Provincial Poverty in the Netherlands"
author: "Timon de Graaf, Mels Dings, Michiel Mast, Sam van Lummel, Stijn, Seppe Naus, Duncan Schröder"
date: "`r Sys.Date()`"
output: pdf_document
---

# Set-up your environment

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

```{r packages}
#install.packages("tidyverse")
#install.packages("giscoR")
#install.packages("sf")
#install.packages("viridis")
#install.packages("readxl")

library(tidyverse)
library(sf)
library(giscoR)
library(viridis)
library(readxl)
```

# Part 1 - Identify a Social Problem

## 1.1 Describe the Social Problem

This project investigates trends in poverty across Dutch provinces from 2011 onward, using CBS microdata on low-income households. We analyze both the absolute number and the relative percentage of people in poverty to determine whether provincial disparities are narrowing or widening. This analysis helps identify where poverty is structurally persistent versus where progress is being made.

Poverty remains a serious concern in the Netherlands. National averages hide the fact that provinces like Groningen, Flevoland, and Limburg face disproportionately high poverty rates. This issue affects children particularly hard. Our project aims to provide clearer insights into both geographic and demographic patterns of poverty.

# Part 2 - Data Sourcing

## 2.1 Load in the data

```{r loading_data}
poverty_data <- read_excel("combined_poverty_data.xlsx")
names(poverty_data)[1] <- "Inkomensgrens"
```

## 2.2 Provide a short summary of the dataset(s)

```{r summary_data}
summary(poverty_data)
head(poverty_data)
```

## 2.3 Describe the type of variables included

- **Inkomensgrens**: Low-income threshold
- **Duur inkomenspositie**: Duration of low-income status
- **Kenmerken van huishoudens**: Type of households
- **Regio's**: Province
- **Perioden**: Year
- **Particuliere huishoudens (x 1 000)**: Number of low-income households
- **Personen (x 1 000)**: Number of people in poverty
- **Minderjarige kinderen (x 1 000)**: Number of children in poverty
- **Relatieve (%)** columns: Percentages of each group in poverty
- **year**: Manually added year variable

# Part 3 - Quantifying

## 3.1 Data Cleaning

```{r cleaning}
poverty_data <- poverty_data %>%
  mutate(across(
    c(`Particuliere huishoudens (x 1 000)`, `Personen (x 1 000)`,
      `Minderjarige kinderen (x 1 000)`, `Particuliere huishoudens relatief (%)`,
      `Personen relatief (%)`, `Minderjarige kinderen relatief (%)`),
    ~ as.numeric(str_replace_all(., ",", "."))
  ))
```

## 3.2 Generate Necessary Variables

```{r new_variables}
poverty_data <- poverty_data %>%
  arrange(`Regio's`, Perioden) %>%
  group_by(`Regio's`) %>%
  mutate(
    poverty_rate_change = `Personen relatief (%)` - lag(`Personen relatief (%)`),
    poverty_ratio_children_to_total = `Minderjarige kinderen relatief (%)` / `Personen relatief (%)`
  ) %>% ungroup()
```

## 3.2a Add Region Variable

```{r region_variable}
poverty_data <- poverty_data %>%
  mutate(region = case_when(
    str_detect(`Regio's`, "Groningen|Fryslân|FryslÃƒÂ¢n|Drenthe") ~ "North",
    str_detect(`Regio's`, "Overijssel|Gelderland|Flevoland") ~ "East",
    str_detect(`Regio's`, "Noord-Holland|Zuid-Holland|Utrecht") ~ "West",
    str_detect(`Regio's`, "Zeeland|Noord-Brabant|Limburg") ~ "South",
    TRUE ~ NA_character_
  ))
```

## 3.3 Visualize Temporal Variation

... (existing code remains unchanged) ...

## 3.5 Visualize Sub-Population Variation by Region

```{r plot_3.5_region}
poverty_data %>%
  filter(Perioden == 2017) %>%
  group_by(region) %>%
  summarise(avg_ratio = mean(poverty_ratio_children_to_total, na.rm = TRUE)) %>%
  ggplot(aes(x = region, y = avg_ratio, fill = region)) +
  geom_col() +
  geom_hline(yintercept = 1, linetype = "dashed") +
  labs(
    title = "Child-to-Total Poverty Ratio by Region (2017)",
    x = "Region",
    y = "Average Ratio"
  ) +
  theme_minimal()
```

## 3.6 Event Analysis – National Trend

```{r plot_3.6_national_trend}
national_trend <- poverty_data %>%
  group_by(Perioden) %>%
  summarise(national_poverty = mean(`Personen relatief (%)`, na.rm = TRUE))

ggplot(national_trend, aes(x = Perioden, y = national_poverty)) +
  geom_line(size = 1.2, color = "darkblue") +
  geom_point(size = 3, color = "darkblue") +
  geom_vline(xintercept = 2014, linetype = "dashed", color = "red") +
  annotate("text", x = 2014, y = max(national_trend$national_poverty), label = "Post-Crisis Period", vjust = -0.5, angle = 90, color = "red") +
  labs(
    title = "National Poverty Rate Over Time",
    subtitle = "Before and after Eurozone Crisis (2014 as breakpoint)",
    x = "Year",
    y = "Average Poverty Rate (%)"
  ) +
  theme_minimal()
```

# Part 4 - Discussion

## 4.1 Discuss Your Findings

The child-to-total poverty ratio highlighted that **families with children are more severely impacted by poverty** than households without children. In our regional analysis, the **South and West** showed the **highest disparities**, with families with children being **up to 20% more likely** to live in poverty than the general population. However, this pattern is **not confined to specific regions**. Across the entire country, children were consistently more likely to live in poverty than adults, underscoring the need for **targeted interventions that support families with children**. Addressing this disproportionate burden is crucial to **preventing transgenerational poverty** and promoting more equitable social outcomes.

# Part 5 - Reproducibility

## 5.1 GitHub Repository Link

[GitHub Repository](https://github.com/FidellC/Armoede-in-Nederland/tree/main)

## 5.2 References

_Same as in final report._
